Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-07-02 paint-contour-tree.tex
interface fixed

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
The best way to think of our procedure is a meteorology analogy. Take an arbitrary maximum $x$,
and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
is a set of contours, given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.

@@

\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of contours,
each containing a join vertex.
\end{claim}

@@

Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
have up-degree $1$, then $\psi$ is the unique $(f(y)+\delta)$-contour
intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
and $\phi$ is dry.

Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
lies in $\Gamma_y$ (which is wet). 
Hence this limit gives a contour of the interface containing the join $w$.
\end{proof}

Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
The wet portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.

We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.

\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
results in two disjoint simplicial complexes, one consisting entirely of dry points.
\end{claim}

@@

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
\end{compactenum}
\end{minipage}}

>>>>

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
The best way to think of our procedure is a meteorology analogy. Take an arbitrary maximum $x$,
and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.

@@

\begin{claim} \label{clm:inter} For any $x$, each connected component of the interface of $\wet(x,\MM)$ contains a join vertex.
\end{claim}

@@

Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
have up-degree $1$, then $\psi$ is the unique $(f(y)+\delta)$-contour
intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
and $\phi$ is dry.
%
% Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
% both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
% lies in $\Gamma_y$ (which is wet). 
% Hence this limit gives a contour of the interface containing the join $w$.
\end{proof}

Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
The wet portion is determined by cutting $F$ along the interface. Since each component of the interface connected subset of a contour, this is equivalent
to locally cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.

We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.

\begin{claim} \label{clm:cut-int} Let $\phi$ be a connected component of the interface. Then $\cut(\MM,\lift(\phi))$
results in two disjoint simplicial complexes, one consisting entirely of dry points.
\end{claim}

@@

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the connected components by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
\end{compactenum}
\end{minipage}}
